\chapter{ Application }
\label{sec:application}

The primary aim of WebAssembly as a distributable for heterogeneous embedded targets is to enable:

\begin{itemize}
  \item Rapid prototyping of signal processing algorithms on a desktop computer
  \item Cooperation between processing modules from multiple authors
\end{itemize}

To demonstrate how WASM provides these affordances, 
I've documented the porting of a guitar effects chain from desktop audio software 
to a system of WASM modules that run both in the browser and on embedded hardware.

\section{ System Architecture }

A system for deploying WASM-based audio processing can be implemented with three major elements:

\begin{itemize}
  \item A \textit{development environment} for the prototyping of WASM audio modules
  \item A \textit{registry} for compiled WASM modules that share a consistent ABI
  \item A \textit{flashing tool} for composing signal graphs of WASM modules
  and deploying them to an embedded target
\end{itemize}

To demonstrate how these elements work in concert, I've documented their use 
in porting of a guitar effects chain from desktop audio software 
to a system of WASM modules that run both in the browser and on embedded hardware.

\section{ Module Prototyping }

The desired effects chain consists of phaser, distortion, and delay modules. 
Each are authored separately with differing code styles, 
but become interoperable when compiled to a consistent WASM ABI via \textit{DaisyOnline}.

The phaser was originally written in C in the Soundspear's Formula development environment, 
a VST plugin that implements a C code editor and compiler. Adapting the phaser to \textit{DaisyOnline} 
simply involved adjusting the return type of the Formula function call to \lstinline{float}. 
See below image.

The distortion module does simple waveshaping using \lstinline{tanh}, 
and was authored from scratch directly in \textit{DaisyOnline}.

The delay module comes from a pre-existing open-source C++ effects library, \textit{Gimmel}. 
To adapt the delay to \textit{DaisyOnline}, all necessary C++ code from Gimmel 
was copied into the single-file interface.

\section{ Module Registry }
To deliver the precompiled WASM modules to a variety of possible targets, 
a centralized registry can be used to host the binaries.
In this work, the registry is implemented as a simple GitHub repository,
available at ???. A more robust registry might implement such features as versioning, 
metadata, and module validation.

\section{ Flashing Tool }